---
title: "Assignment 09: Data Scraping"
author: "Danlei"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1

#check working directory
getwd()

#loading necessary packages

library(lubridate)
library(tidyverse)
library(rvest)

#setting theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
LWSP.webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')
LWSP.webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3

#scraping for water system name
water.system.name <- LWSP.webpage %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
water.system.name

#scraping for PWSID
pwsid <- LWSP.webpage %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
pwsid
  
ownership <- LWSP.webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
ownership
  
#scraping for max daily use
max.withdrawals.mgd <- LWSP.webpage %>%
  html_nodes("th~ td+ td") %>%
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4

#creating new month dataframe
month <- c(1,5,9,2,6,10,3,7,11,4,8,12)

#creating dataframe
df.dailywithdrawals <- data.frame("Month" = as.numeric(month),
                                  "Year" = rep(2021,12),
                                  "Ownership" = as.character(ownership),
                                  "Water System Name" = as.character(water.system.name),
                                  "PWSID" = as.character(pwsid),
                                  "Max_Withdrawals_MGD" = as.numeric(max.withdrawals.mgd))
df.dailywithdrawals <- df.dailywithdrawals %>%
  mutate(Date = my(paste(Month, "-", Year)))

#5

#creating line plot of max daily withdrawals across months for 2021 in Durham
ggplot(df.dailywithdrawals, aes(x = Date, y = Max_Withdrawals_MGD)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = paste("2021 water usage data for Durham"),
       y = "Max Withdrawals (mgd)",
       x = "Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.

#create scraping function
year <- 2021

scrape.it <- function(the_year, the_pwsid){
  
  #retrieving website contents
  the_website <- read_html(paste0
                           ('https://www.ncwater.org/WUDC/app/LWSP/report.php?', 'pwsid=', the_pwsid, '&year=', the_year))
  
  #scraping data items
  water.system.name <- the_website %>%
    html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  pwsid <- the_website %>%
    html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
  ownership <- the_website %>%
    html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
  max.withdrawals.mgd <- the_website %>%
      html_nodes('th~ td+ td') %>% html_text()
  
  #converting to dataframe
  df.dailywithdrawals2 <- data.frame("Month" = as.numeric(month),
                                     "Year" = rep(the_year, 12),
                                     "Ownership" = as.character(ownership),
                                     "Water System Name" = as.character(water.system.name),
                                     "PWSID" = as.character(pwsid),
                                     "Max_Withdrawals_MGD" = as.numeric(max.withdrawals.mgd)) %>%
    mutate(Date = my(paste(Month, "-", Year)))
  
  Sys.sleep(1)
  
  return(df.dailywithdrawals2)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7

#extracting 2015 data for Durham
the_year <- 2015
the_pwsid <- as.character('03-32-010')

#assigning to dataframe
the_dataframe <- data.frame(scrape.it(the_year, the_pwsid))
print(the_dataframe)

#plotting max daily withdrawals across months for 2015 in Durham
ggplot(the_dataframe, aes(x = Date, y = Max_Withdrawals_MGD)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = paste("2015 water usage data for Durham"),
       y = "Max Withdrawal (mgd)",
       x = "Date")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8

#extracting 2015 data for Asheville
the_year <- 2015
the_pwsid <- as.character('01-11-010')

#assigning to dataframe
the_dataframe_asheville <- data.frame(scrape.it(the_year, the_pwsid))
print(the_dataframe_asheville)

#creating joined dataframe of Asheville and Durham 2015 data
withdrawals_joined <- merge(x = the_dataframe,
                            y = the_dataframe_asheville,
                            by = c("Date", "Month", "Year"))

#creating plot of joined dataframe
ggplot_joined <- withdrawals_joined %>%
  gather(WaterSystem, City, Max_Withdrawals_MGD.x, Max_Withdrawals_MGD.y) %>%
  ggplot(aes(x = Date, y = City, colour = WaterSystem)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  scale_shape_discrete(labels = c("Durham", "Asheville")) +
  scale_colour_discrete(labels = c("Durham", "Asheville")) +
  labs(title = paste("2015 water usage data for Durham and Asheville"),
       y = "Max Withdrawal (mgd)",
       x = "Date")

print(ggplot_joined)
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9

#extracting 2010-2019 data for Asheville
the_years <- rep(2010:2019)
the_pwsid <- as.character('01-11-010')

#applying scrape function
asheville_df2 <- lapply(X = the_years,
                        FUN = scrape.it,
                        the_pwsid = the_pwsid)

#conflate into single dataframe
asheville_df_20102019 <- bind_rows(asheville_df2)

#creating a plot of  daily max withdrawals from 2010-2019 in Asheville
ggplot(asheville_df_20102019, aes(x = Date, y = Max_Withdrawals_MGD)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = paste("2010 - 2019 water usage data for Asheville"),
       y = "Max Withdrawal (mgd)",
       x = "Date")


```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

>ANSWER: Yes, Asheville has a trend in water usage over time by looking at the plot. The plot shows an increase in water usage from 2015 to 2019.