---
title: "Assignment 2: Coding Basics"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.


```{r}
#1. 
q1_seq <- seq(1, 100, by = 4)

#2. 
#create vector
x <- c(q1_seq)

#find mean
result.mean <- mean(x)
print(result.mean)

#3. 
#find median
result.median <- median(x)
print(result.median)
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5 and #6.

vector1 <- c('Marianne', 'Felix', 'Ingrid', 'Claude') #character vector
vector1
vector2 <- c(90, 60, 45, 70) #numeric vector
vector2
vector3 <- c(TRUE, TRUE, FALSE, TRUE) #logical vector
vector3

#7. Combining vectors into data frame

dataframe1 <- data.frame(vector1,vector2,vector3)
names(dataframe1) <- c("Name", "Score", "Passed"); View(dataframe1)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: In a matrix, all of the elements have to be the same type of data but in a data frame the columns can have different types of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#vector of test scores
vector1 <- c(33, 40, 80, 73)

#ifelse function to determine passing
ifelse(vector1 < 50, "FAIL", "PASS")

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'if' and 'else' didn't work because if the test expression failed, then nothing would get executed. 'ifelse' worked because it presented parameters of what would pass, but also what should be returned in case of a fail.
