---
title: "Assignment 4: Data Wrangling"
author: "Danlei Zou"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct7th @ 5:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1

#checking working directory and loading lubridate and tidyverse
getwd()
library(lubridate)
library(tidyverse)

#loading the four EPA Air raw datasets with stringAsFactors
O3NC2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)
O3NC2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
PM25NC2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)
PM25NC2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)

#2

#exploring dimensions of datasets
dim(O3NC2018)
dim(O3NC2019)
dim(PM25NC2018)
dim(PM25NC2019)

#exploring column names of datasets
colnames(O3NC2018)
colnames(O3NC2019)
colnames(PM25NC2018)
colnames(PM25NC2019)

#exploring structure of datasets
str(O3NC2018)
str(O3NC2019)
str(PM25NC2018)
str(PM25NC2019)


```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3

#changing date to date format in datasets

O3NC2018$Date <- mdy(O3NC2018$Date)
O3NC2019$Date <- mdy(O3NC2019$Date)
PM25NC2018$Date <- mdy(PM25NC2018$Date)
PM25NC2019$Date <- mdy(PM25NC2019$Date)

#checking to make sure date is now in date format
class(O3NC2018$Date)

#4

#selecting columns for each dataset

O3NC2018.subset <- select(O3NC2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

O3NC2019.subset <- select(O3NC2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

PM25NC2018.subset <- select(PM25NC2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

PM25NC2019.subset <- select(PM25NC2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#checking dimensions of new subsets to make sure there are 7 columns in each
dim(O3NC2018.subset)
dim(O3NC2019.subset)
dim(PM25NC2018.subset)
dim(PM25NC2019.subset)

#5

#filling cells in AQS_PARAMETER_DESC to "PM2.5"
PM25NC2018.subset$AQS_PARAMETER_DESC = "PM2.5"

PM25NC2019.subset$AQS_PARAMETER_DESC = "PM2.5"

#checking length of columns in PM2.5 datasets after filling 
view(PM25NC2018.subset$AQS_PARAMETER_DESC)
length(PM25NC2018.subset$AQS_PARAMETER_DESC)

view(PM25NC2019.subset$AQS_PARAMETER_DESC)
length((PM25NC2019.subset$AQS_PARAMETER_DESC))


#6

#saving processed datasets to Data/Processed folder
write.csv(O3NC2018.subset, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(O3NC2019.subset, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(PM25NC2018.subset, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(PM25NC2019.subset, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7

#combining processed EPA air datasets
EPAairCombined <- rbind(O3NC2018.subset, O3NC2019.subset, PM25NC2018.subset, PM25NC2019.subset)

#checking dimensions of combined dataset
dim(EPAairCombined)

#8

#wrangling dataset with pipe function to satisfy the conditions given
EPAairCombined2 <-
  EPAairCombined %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQIValue = mean(DAILY_AQI_VALUE),
            meanLatitude = mean(SITE_LATITUDE),
            meanLongitude = mean(SITE_LONGITUDE)) %>%
  mutate(month = month(Date), year = year(Date))

#checking summary and dimensions of new dataset after pipe function
summary(EPAairCombined2)

dim(EPAairCombined2)

#9

#spreading ozone and PM2.5 AQI values into separate columns
EPAairCombined.spread <- pivot_wider(EPAairCombined2, names_from = AQS_PARAMETER_DESC, values_from = meanAQIValue)

#10

#calling up dimensions of spread dataset
dim(EPAairCombined.spread)

#11

#saving processed dataset to Data/Processed folder
write.csv(EPAairCombined.spread, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a and 12b

#using split-apply-combine to generate summary dataset with meanAQI values of ozone and PM25
EPAairCombined.spread.summary <-
  EPAairCombined.spread %>%
  group_by(Site.Name, month, year) %>%
  summarise(meanAQIozone = mean(Ozone),
            meanAQIPM25 = mean(PM2.5)) %>%
  #pipe function to remove instances without month and year
  drop_na(meanAQIozone, meanAQIPM25)

#13

#checking dimensions of summary dataset
dim(EPAairCombined.spread.summary)
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: The drop_na function will get rid of the rows with NAs, whereas the na.omit function will delete instances of NAs from the dataframe.
